---

title: "Modèles matriciels et applications"
group: "Rudy, Tanguy, Manon, Garan, Anaëlle, Oceane, Victor" 

bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

Le lien du site : https://zenodo.org/records/8251495 


Pistes 

Calcul 

Taux de succès 
Taux de survie

Oceane : ANOVA pour voir effet de l'habitat sur les variables 

voir si habitat préférentiel, région,…
voir quelles facteurs expliquent taux de succès

voir si décalages entre dates de départ du nid -> facteurs expliquant 

voir si habitat préferentiel 

MAIS ATTENTION à ne pas être biaisé : du genre que des oiseaux d’une certaines espèces ne soit présents que dans un habitat

refaire analyse de bases cf ASA 

# Document R modifié et avec ajouts

## Import des données

**Do not forget to set your working directory in the right place**
**Menu : Session/Set Working Directory/Choose directory**
*And indicate Git/2024_MODE_OCR*


```{r}
# Datatset import under the name of "data"
data <- read.csv("Bird_dataset.csv", sep=";", header = TRUE)
```

```{r}
# Charge the librairies
library(ggplot2)
```

```{r}
# View the firsts lignes of the dataset
head(data)
```
## Data exploration
# Repartition of the data through time

```{r}
# Creation of a barplot
ggplot(data, aes(x="Year"))+
  geom_bar(stat="bin", width=0.7, fill="steelblue")+
  theme_minimal()
```



I. ANOVA (Oceane  G.)

Performing an ANOVA to see the effect of habitat on the variables in the Bird_dataset.csv dataset

ANOVA : 
- ANOVA is a statistical method that simultaneously compares means across several groups to determine if observed differences are due to chance or reflect genuine distinctions.
- A one-way ANOVA uses one independent variable. A two-way ANOVA uses two independent variables.
- By partitioning total variance into components, ANOVA unravels relationships between variables and identifies true sources of variation.
- ANOVA can handle multiple factors and their interactions, providing a robust way to better understand intricate relationships.






