---

title: "Modèles matriciels et applications"
group: "Rudy, Tanguy, Manon, Garan, Anaëlle, Oceane, Victor" 

bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

Le lien du site : https://zenodo.org/records/8251495 


Pistes 

Calcul 

Taux de succès 
Taux de survie

voir si habitat préférentiel, région,…
voir quelles facteurs expliquent taux de succès

voir si décalages entre dates de départ du nid -> facteurs expliquant 

voir si habitat préferentiel 

MAIS ATTENTION à ne pas être biaisé : du genre que des oiseaux d’une certaines espèces ne soit présents que dans un habitat

refaire analyse de bases cf ASA 

# Document R modifié et avec ajouts

## Import des données

**Do not forget to set your working directory in the right place**
**Menu : Session/Set Working Directory/Choose directory**
*And indicate Git/2024_MODE_OCR*


```{r}
# Datatset import under the name of "data"
data <- read.table("Bird_dataset.csv", sep="\t", header = TRUE)
```

```{r}
# Charge the librairies
library(ggplot2)
```

```{r}
# View the firsts lignes of the dataset
head(data)
```
## Data exploration
# Repartition of the data through time

```{r}
# Creation of a barplot
ggplot(data, aes(x="Year"))+
  geom_bar(stat="bin", width=0.7, fill="steelblue")+
  theme_minimal()
```



I. ANOVA (Oceane  G.)


Analysis of variance (ANOVA) is one of the most widely used statistical methods in various disciplines. It allows to model the relationship between a quantitative dependent variable $Y$ according to the different modalities of one or more qualitative explanatory variables $Xs$.

We will performing an ANOVA to see the effect of qualitatives variables (state, habitat, species, year) on the quantitatives (initiation_date, number_fledged, nest_end_date) variables in the Bird_dataset.csv dataset.

  
  A. Effect on Initiation_date
  
  
  
  B. Effect on Number_fledged



  C. Effect on Nest_end_date

