---

title: "Modèles matriciels et applications"
group: "Rudy, Tanguy, Manon, Garan, Anaëlle, Oceane, Victor" 

bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

# Ideas of modelisation : 

Le lien du site : https://zenodo.org/records/8251495 

- Océane : ANOVA

- Manon : ACP

- Rudy : Number_Fledged





Pistes 

Calcul 

Taux de succès 
Taux de survie

voir si habitat préférentiel, région,…
voir quelles facteurs expliquent taux de succès

voir si décalages entre dates de départ du nid -> facteurs expliquant 

voir si habitat préferentiel 

MAIS ATTENTION à ne pas être biaisé : du genre que des oiseaux d’une certaines espèces ne soit présents que dans un habitat

refaire analyse de bases cf ASA 







# Document R modifié et avec ajouts

## Import des données

**Do not forget to set your working directory in the right place**
**Menu : Session/Set Working Directory/Choose directory**
*And indicate Git/2024_MODE_OCR*


```{r}
# Datatset import under the name of "data"
# data <- read.table("Bird_dataset.csv", sep="\t", header = TRUE)
bird <- read.table("Bird_dataset.txt", header = TRUE, stringsAsFactors = TRUE)

data <- bird
```

```{r}
# Charge the librairies
library(ggplot2)
```

```{r}
# View the firsts lignes of the dataset
head(data)
```
## Data exploration
# Repartition of the data through time



# Rudy part : 

We can use the number of fledged birds as response variable and try to explain this by several variables. We can analyse if the number of fledged birds depends on the species. 

H0 : there is no differences between species

First, we observe the data with a boxplot
```{r}
# Compari
ggplot(data, aes(x = Species, 
                 y = Number_Fledged
                 )
       ) +
  geom_boxplot(outlier.shape = 0.2)
                 
                 
```

Then, we test the normality of the data : 

```{r}
sp_fledged <- data |>
  dplyr::select(
    Species, 
    Number_Fledged
    ) |>
  dplyr::group_by(Species) |>
  dplyr::summarise()
  

sp_fledged
```














