---

title: "Modèles matriciels et applications"

group: "Rudy, Tanguy, Manon, Garan, Anaëlle, Oceane, Victor" 


bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

Le lien du site : https://zenodo.org/records/8251495 


Pistes 

Calcul 

Taux de succès 
Taux de survie

voir si habitat préférentiel, région, …
voir quelles facteurs expliquent taux de succès

voir si décalages entre dates de départ du nid -> facteurs expliquant 

voir si habitat préferentiel 

MAIS ATTENTION à ne pas être biaisé : du genre que des oiseaux d’une certaines espèces ne soit présents que dans un habitat

refaire analyse de bases cf ASA 

# Document R modifié et avec ajouts


## Import des données

**Do not forget to set your working directory in the right place**
**Menu : Session/Set Working Directory/Choose directory**
*And indicate Git/2024_MODE_OCR*


```{r}
# Dataset import under the name of "data"
data <- read.csv("Bird_dataset.csv", sep=";", header = TRUE, stringsAsFactors = TRUE)
data$Nest_Fate <- as.factor(data$Nest_Fate)
data$Year <- as.factor(data$Year)

# Checking if there are missing data
colSums(is.na(data))
```

```{r}
# Charge the librairies
library(ggplot2)
```

```{r}
# View the firsts lines of the dataset
head(data)
```
# Variables presentation:

- *Nest_ID* : the code to identify each nest.
- *State* : the state in which the nest has been found. 2 levels factor : Illinois and Wisconsin.
- *County* : the county in the state where the state has been found. 14 levels factor.
- *Habitat* : the type of habitat in which the nest has been found. 4 levels factor.
- *Species* : the species of the bird that has build the nest. 8 levels factor.
- *Year* : the year of the nest building. 21 levels factor.
- *Nest_Fate* : variable that says if there is a least one bird hatch in the nest (1) or not (0).
- *Initiation_Date* : the date when the bird hatch from the 1st of April (1), May the 1st (31), June the 1st (62), etc.
- *Number_Fledged* : the number of birds that hatch on the nest when they hatch. Quantitative variable.
- *Nest_End_Date* : the date where the nest either failed of the birds successfully hatch.

We want to study the impact of four qualitative variables, the habitat, the county, the species and the year over the number of young hatched in the nest. 


## Data exploration

### Outlayer values in the response variable $Y$ (the number of young in the nest, Number_Fledged) and values distribution:

```{r datahist, include=TRUE, fig.height=5, fig.width=5}
par(mfrow=c(2,2))
# Boxplot
boxplot(data$Number_Fledged,col='dodgerblue3',ylab='Number of young in nest')
# Cleveland plot
dotchart(data$Number_Fledged,pch=16,col='dodgerblue3',xlab='Number of young in nest')
# Histogram
hist(data$Number_Fledged,col='dodgerblue3',xlab="Number of young in nest",main="")
# Quantile-Quantile plot
qqnorm(data$Number_Fledged,pch=16,col='dodgerblue3',xlab='')
qqline(data$Number_Fledged,col='red',lwd=2)
```

There isn't any outlayer values.
There are a lot of value at 0. Maybe we can study only the number of young when there are some. There could be a normal distribution.

```{r datafact, include=TRUE}
# Retirer les lignes où une valeur égale à 0 est présente dans n'importe quelle colonne
data_filtered <- subset(data, Number_Fledged != 0)
```

```{r datahist, include=TRUE, fig.height=5, fig.width=5}
par(mfrow=c(2,2))
# Boxplot
boxplot(data_filtered$Number_Fledged,col='dodgerblue3',ylab='Number of young in nest')
# Cleveland plot
dotchart(data_filtered$Number_Fledged,pch=16,col='dodgerblue3',xlab='Number of young in nest')
# Histogram
hist(data_filtered$Number_Fledged,col='dodgerblue3',xlab="Number of young in nest",main="")
# Quantile-Quantile plot
qqnorm(data_filtered$Number_Fledged,pch=16,col='dodgerblue3',xlab='')
qqline(data_filtered$Number_Fledged,col='red',lwd=2)
```

Now we have a better view of the data.


### Outlayer values in the explanatory variable $Xs$ and values distribution:

```{r datafact, include=TRUE}
# Factor County
summary(data_filtered$County)
# Factor Habitat
summary(data_filtered$Habitat)
# Factor Species
summary(data_filtered$Species)
# Factor Year
summary(data_filtered$Year)
```
Problem: there aren't the same number of observations in each level of each factors.
Suggestion: normalize by the number of observation for each level.


### Analyze the relation between Y and Xs

```{r datagraph, include=TRUE, fig.height=7, fig.width=7}

par(mfrow=c(2,2))

# County
boxplot(Number_Fledged~County, varwidth = TRUE, ylab = "Number of young in nest", xlab = "County", col=terrain.colors(14), main = "", data=data_filtered)

# Habitat
boxplot(Number_Fledged~Habitat, varwidth = TRUE, ylab = "Number of young in nest", xlab = "Habitat", col=terrain.colors(4), main = "",data=data_filtered)

# Species
boxplot(Number_Fledged~Species, varwidth = TRUE, ylab = "Number of young in nest", xlab = "Species", col=terrain.colors(8), main = "",data=data_filtered)

# Year
boxplot(Number_Fledged~Year, varwidth = TRUE, ylab = "Number of young in nest", xlab = "Year", col=terrain.colors(21), main = "",data=data_filtered)
```
It's not easy to see if the variables have an impact because some of them have a lot of levels, but it is possible to see differences between them. The analysis will tell us if they really got an imapct over the number of young in the nest.


### Analyze the interactions between Xs

There are a lot of levels so it's hard to plot the interactions. We will see if there are some in the analysis.

```{r dataInter, include=TRUE, fig.height=3, fig.width=8}

par(mfrow=c(1,3))
# Interactions between County & Habitat
plot(Number_Fledged~County,type='n',ylab = "Number of young in nest",xlab="County", data=data_filtered)
points(Number_Fledged[Habitat=="Pasture"]~County[Habitat=="Pasture"],pch=16,cex=2,col='#000099', data=data_filtered)
points(Number_Fledged[Habitat=="Cool-season Grassland"]~County[Habitat=="Cool-season Grassland"],pch=16,cex=2,col='green', data=data_filtered)
points(Number_Fledged[Habitat=="Prairie"]~County[Habitat=="Prairie"],pch=16,cex=2,col='red', data=data_filtered)
points(Number_Fledged[Habitat=="Warm-season Grassland"]~County[Habitat=="Warm-season Grassland"],pch=16,cex=2,col='yellow', data=data_filtered)




```


# Checking collinearity between categorical independent variables Xs:

```{r datacolin, include=TRUE, fig.height=4, fig.width=8}

contingency_table<-table(data_filtered$Habitat,data_filtered$Species)
print(contingency_table)

chi2_test<-chisq.test(contingency_table)
print(chi2_test)
print(chi2_test$expected)


```
Problem : for most of the variables the chi² test can't be done because of the low number of values in each group. The only one that fulfill the conditions and therefore can be made is between Habitat and Species but the result shows that the variables are correlated.


### Statistical Analysis

# Building of the model



```{r fullmodel,include=TRUE}
# Model formulation
mod1<-glm(Number_Fledged~ County
        + Species
        + Habitat
        + Year
        ,data=data_filtered
        ,family=poisson(link="log"))
#type de la variable : poisson et fonction de lien de type log
# Then we check for significance
drop1(mod1,test="Chi")
```
We decide to remove the less significative variable : here it's Year.

```{r fullmodel,include=TRUE}
# Model formulation
mod1<-glm(Number_Fledged~ County
        + Species
        + Habitat
        ,data=data_filtered
        ,family=poisson(link="log"))
#type de la variable : poisson et fonction de lien de type log
# Then we check for significance
drop1(mod1,test="Chi")
```
Now the less significative is Habitat

```{r fullmodel,include=TRUE}
# Model formulation
mod1<-glm(Number_Fledged~ County
        + Species
        ,data=data_filtered
        ,family=poisson(link="log"))
#type de la variable : poisson et fonction de lien de type log
# Then we check for significance
drop1(mod1,test="Chi")
```
County is still not significative so we have to remove it.


```{r fullmodel,include=TRUE}
# Model formulation
mod1<-glm(Number_Fledged~ Species
        ,data=data_filtered
        ,family=poisson(link="log"))
#type de la variable : poisson et fonction de lien de type log
# Then we check for significance
drop1(mod1,test="Chi")

summary(mod1)
```
We have the final model with only the Species.
 
- $Species{Bobolink }$ = 0 (intercept)  
- $Species{Dickcissel }$ = $-0.18^{**}$ 
- $Species{Eastern Meadowlark }$ = $-0.14^{**}$ 
- $Species{Grasshopper Sparrow }$ = $-0.05$ 
- $Species{Henslow's Sparrow }$ = $-0.18$ 
- $Species{Savannah Sparrow }$ = $-0.19^{***}$ 
- $Species{Vesper Sparrow }$ = $-0.32^{***}$
- $Species{Western Meadowlark }$ = $-0.04$

The equation of the model is the following: 

$$ log(Species\:Richness) = 1.29^{***} + (Species{Bobolink} = 0 ;\:Species{Dickcissel} = -0.18^{**} ; \:Species{Eastern Meadowlark} = -0.14^{**} ; \:Species{Grasshopper Sparrow} = -0.05 ; \:Species{Henslow's Sparrow} = -0.18 ; \:Species{Savannah Sparrow = -0.19^{***} ; \:Species{Vesper Sparrow = -0.32^{***} ; \:Species{Western Meadowlark} = -0.04)$$
