---
title: "Effets du pâturage sur les prairies suédoises"
author: "Bastien, Elia, Caroline, Alexia, Marine, Lily, Julie" 
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      INTRODUCTION
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Terrestrial ecosystems were dominated for several million years by large herbivorous mammals, which then declined at the end of the Pleistocene under the pressure of the first prehistoric human populations. The loss of these large herbivore species (aurochs, wild horses and anes, woolly rhinoceros...) was also synonymous with the loss of the ecosystem services they provided. @lundgren2020human

For several years, the “rewilding” movement has been aiming to restore these ecosystem services by reintroducing species into large European natural areas, and the present study seeks to assess the extent to which introduced herbivore species have an influence on vegetation. @garido2019experimental

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      MATERIALS AND METHODS
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


The study site is located approximately 17 km south of Uppsala, Sweden. On this study site, three plots were used. Each plot was divided in two, with an ungrazed part and a grazed part. Four young Gotland stallions were placed in each grazing area for three years.

Numerous variables were studied:
- plant height at maturity
- seed mass
- species richness
- plot area occupied by anemogamous plants 
- plot area occupied by entomogamous plants 
- RCS score *.
Other information was noted at each survey, such as the season, year and observer who carried out the inventory. Information on the insects found on the plot was also recorded, but will not be used in our analysis.

* The RCS score corresponds to an index of the strategy used by the plants. @Wood2018microbiomes

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      RESEARCH QUESTIONS
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

We have chosen to use the  @garido2019experimental dataset to answer three different questions. Firstly, does grazing have an effect on the plant species richness of plots? Secondly, does grazing have an effect on plant morphology? And finally, does grazing have an effect on plant pollination strategies?

To answer these questions, we transformed the initial dataset to obtain a dataset of 404 records. The variables used are :
- treatment (grazed or ungrazed)
- season
- year
- cover.area.bee (plot area occupied by entomogamous plants)
- cover.area.gene (plot area occupied by anemogamous plants)
- logh (log of plant height)
- logS (log of seed mass)
- richness (species richness)
- observer
- R.score
- S.score
- C.score
- ratio poll (cover.area.bee/cover.area.gene)

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      DATA IMPORT
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


```{r}
df <- read.table('pollen.csv', sep = ';',stringsAsFactors=TRUE, header = TRUE)
df$year <- as.factor(df$year)
summary (df)
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      PACKAGES IMPORT
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


```{r}

library(tidyverse)
library(FactoMineR)
library(dplyr)
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      PARTIE 1 : Effet du paturage sur la richesse specifique
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Visualisation des donnees
```{r}
ggplot(data=df)+
  geom_boxplot(aes(x=treatment, y=num.species))+
  geom_jitter(aes(x=treatment, y=num.species)) + 
  labs(x="Treatment", y="Species richness") +
  ggtitle('Effect of grassland on species richness')
```
La richesse specifique des zones paturees semble etre plus importante que dans les zones non paturees. 

```{r}
hyp1<-lm(num.species~1,data=df)
hyp0<-lm(num.species~treatment,data=df)
anova(hyp0,hyp1)
```
La p-value est inferieure a 0,05 donc le paturage a un effet significatif sur la richesse specifique. 

# Choix du modele : dans cette partie, nous considerons que le paturage (treatment) depend de la saison et de l'annee d'observation. 

```{r}
modComplet<-LinearModel(num.species~(treatment+season+year)*(treatment),data=df,selection = "aic")
modComplet

```
Les resultats pour le modele complet sont les suivants : p-value: 2.554e-16, AIC = 649, BIC = 681
Cependant, le modele choisi selon le critere AIC montre les resultats suivants : p-value: 1.41e-17, 
AIC = 645.2 et BIC = 669.2. Le critere AIC etant plus petit pour ce modele que le modele complet, on choisit
de conserver : num.species ~ treatment + season + year + treatment:year. Il y a donc interaction entre le paturage et son annee d'observation.   

# Validation du modele choisi avec le test de Bartlett
```{r}
modelAIC<-lm(num.species~treatment+season+year+treatment:year,data=df)

res<-residuals(modelAIC)
boxplot(res~treatment,data=df)
bartlett.test(res~treatment,data=df)

boxplot(res~season,data=df)
bartlett.test(res~season,data=df)

boxplot(res~year,data=df)
bartlett.test(res~year,data=df)

boxplot(res~treatment:year,data=df)
df$interaction <- interaction(df$treatment, df$year)
bartlett.test(res ~ interaction, data = df)
```
Pour les tests de Bartlett, les p-value sont toutes supérieures à 0.05, ce qui veut dire que l'homoscédasticité est respectée pour tous nos parametres soit que les variances sont toutes egales pour toutes nos observations.

#Test normalité des résidus avec Shapiro.
hist(res,main="Histogramme residus",xlab="Residus")
shapiro.test(res)

La normalite des residus est respectee car notre p-value est superieure a 0,05 : les donnees sont compatibles avec une distribution normale. Notre modele lineaire est donc valide car la normalite des residus est une hypothese cle pour appliquer ce modele.

# Conclusion partie 1 sur la richesse spécifique
La richesse spécifique depend donc du paturage, de la saison et de l’annee, ainsi que de l’interaction
paturage:annee