---
title: "Effets du pâturage sur les prairies suédoises"
author: "Bastien, Elia, Caroline, Alexia, Marine, Lily, Julie" 
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---


PARTIE 3: GRAZINF EFFECT ON POLLINISATION STRATEGIES 



```{r}
# renommer la premiere colonne pour Marine et Alexia car var treatment est ?..treatment et pose pblm
names(df)[1] <- "treatment"
names(df)
```


```{r}
# Boxplot 
par(mfrow=c(1,2))
boxplot(df$RatioPoll~df$treatment, varwidth = TRUE, ylab = "Ratio Pollinisation", xlab = "treatment", col='blue', main = "")

```

The boxplot shows no significant differences between plots grazed and those not grazed.

```{r}
# Anova
poll1<-lm(RatioPoll~1,data=df)
poll2<-lm(RatioPoll~treatment,data=df)
anova(poll1,poll2)
```

The P-value is not significant at the 5% threshold, so we reject the hypothesis that grazing has an effect on the RatioPoll. The Anova test confirms our graphical observation: there is no difference between the grazed and natural plots.

```{r}
poll3<-LinearModel(RatioPoll~(treatment+season+year+observer+R.score+S.score+C.score)*
                     (treatment+season+year+observer+R.score+S.score+C.score),data=df,selection = "aic")
poll3
```

The model proposed by R does have a lower BIC than our initial model. However, some of the interactions retained by R have no significant effect. We decided to test a model without these interactions. 

```{r}
poll4<-LinearModel(RatioPoll~( treatment + season + year + 
    observer + R.score + S.score + C.score + treatment:R.score + 
    treatment:C.score + year:C.score + 
    observer:R.score + R.score:C.score),data=df)
poll4
```
We have observed that the R.score:C.score interaction is not significant, we test a new model without it.
```{r}
poll5<-LinearModel(RatioPoll~( treatment + season + year + 
    observer + R.score + S.score + C.score + treatment:R.score + 
    treatment:C.score + year:C.score + 
    observer:R.score ),data=df)
poll5
```
We delete R.score because it is the less significant.

```{r}
poll6<-LinearModel(RatioPoll~( season + C.score + treatment:R.score + 
    treatment:C.score + year:C.score ),data=df)
poll6
```
