---

title: "Effets du pâturage sur les prairies suédoises"
group: "Bastien, Elia, Caroline, Alexia, Marine, Lily, Julie" 
>>>>>>> 75e9da09555ae3bad45e7f54a26d7c51188f42d7

bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---
IMPORTATION DES PACKAGES

```{r}
library(tidyverse)
library(FactoMineR)
library(dplyr)
```

IMPORTATION DES DONNEES

```{r}
df <- read.table("pollen.csv", sep = ";" , header = TRUE)
```

INTRODUCTION 

Les ecosystemes terrestres ont ete domines pendant plusieurs millions d'annees par les grands mammiferes herbivores, qui ont ensuite decline à la fin du Pleistocene sous la pression des premieres populations humaines prehistoriques. La perte de ces especes de grands herbivores (aurochs, chevaux et anes sauvages, rhinoceros laineux...) a egalement ete synonyme de la perte des services ecosystemiques qu'ils apportaient. (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7148574/)

Depuis plusieurs annees, le courant du reensauvagement ("rewilding" en anglais) a pour but de restaurer ces services ecosytemique en reintroduisant des especes dans les grands espaces naturels europeens.L'etude presente cherche a evaluer dans quelle mesure les especes d'herbivores introduites ont une influence sur la vegetation. (https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2664.13338)


PRESENTATION DE L'EXPERIENCE

Blabla sur l'experience





# Part 2 : effect of grazing on the morphology 

In this part, we'll try to analyse the effect of treatment (grazing), a qualitative variable, on the morphological parameters of the plants (height and seed mass), quantitative variables. 

### **1. Effect of grazing on the plant height**

[**Potential relationship between Height and effect of treatment**]{.underline} **:**

Let's start by representing plant size as a function of the applied treatment (grazing or no grazing) to have an idea of the possible effect.

```{r effect_treatment_on_height_graph}
boxplot(df$logHeight.score~df$treatment, varwidth = TRUE, ylab = "plant height (log)", xlab = "Treatment", col='darkgrey', main = "", title = "effect of grazing on the height")

ggplot(data=df)+
  geom_boxplot(aes(x=treatment, y=logHeight.score))+
  geom_jitter(aes(x=treatment, y=logHeight.score))+
  ylab("Plant height (log)") +
  xlab("Treatment")

```
Interpretation :

The median for the no grazing treatment is slightly higher than for the grazing treatment. The data also seems slightly more dispersed and spread out.
So we could say that the plants without grazing are potentially higher than the ones which had the treatment.

[**Statistical analysis of the relationship between Height and effect of treatment**]{.underline} **:**

We saw earlier that there could be a potential effect of treatment on the plant's height, we now want to know if this effect is really significant.
 
We perform an ANOVA to compare a random model with our model.

Significance test:
Hypothesis H0: Grazing has no effect on plant size.
Hypothesis H1: Grazing has an effect on plant size.


```{r model_treatment_height_anova}
hyp0<-lm(logHeight.score~1,data=df)
hyp1<-lm(logHeight.score~treatment,data=df)
anova(hyp0,hyp1)
```

The p-value is 2.39e-4 < 0.05. The test is significant, so hypothesis H0 is rejected. Grazing significantly reduces plant size.


### **2. Effect of grazing on the seed mass**

Now, let's study the seed mass of plants in relation to grazing.

[**Potential relationship between seed mass and effect of treatment**]{.underline} **:**

Let's start by representing the seed mass of plants as a function of the applied treatment (grazing or no grazing) to have an idea of the possible effect.

```{r effect_treatment_on_seed_mass_graph}
boxplot(df$logSeedMass.score~df$treatment, varwidth = TRUE, ylab = "Seed mass (log)", xlab = "Treatment", col='darkgrey', main = "", title = "effect of grazing on seed mass")

ggplot(data=df)+
  geom_boxplot(aes(x=treatment, y=logSeedMass.score))+
  geom_jitter(aes(x=treatment, y=logSeedMass.score))+
  labs(y = "Seed mass (log)", x = "Treatment")
```
Interpretation :

The medians are almost identical. The data for the no grazing treatment also seems more dispersed and spread out. So we could say that there is potentially no effect of treatment on seed mass.

[**Statistical analysis of the relationship between Height and effect of treatment**]{.underline} **:**

We saw earlier that there could be no potential effect of treatment on seed mass. We now want to know if this is really significant.
 
We perform an ANOVA to compare the model where we only consider the effect of grazing with a model where the effect of grazing is random.

Significance test:
Hypothesis H0: Grazing has no effect on seed mass.
Hypothesis H1: Grazing has an effect on seed mass.



```{r model_treatment_seed_mass_anova}
hyp2<-lm(logSeedMass.score~1,data=df)
hyp3<-lm(logSeedMass.score~treatment,data=df)
anova(hyp2,hyp3)
```
The p-value = 0.3573 > 0.05, so the seed mass under hypothesis H1 is not significantly different from that under hypothesis H0.
We can conclude that there is no effect of grazing on seed mass.

### **3. Is the height of plants only affected by grazing ?**

We will now analyze the effect of grazing on plant size more precisely. We want to know if the size is influenced by grazing (treatment) alone or by its interaction with other factors.

For this analysis, we will test the effects of grazing and its interaction with other factors on plant size through model analysis using the BIC criterion.

```{r final_linear_model_height_bic}
hyp4<-LinearModel(logHeight.score~(treatment+season+year+observer+logSeedMass.score+RatioPoll+R.score+C.score+S.score)*
                    (treatment+season+year+observer+logSeedMass.score+RatioPoll+R.score+C.score+S.score), data=df, selection = "bic")
hyp4

```

After the analysis, the final model with the lower BIC is : logH ~ treatment + logS + RatioPoll + R.score + C.score + S.score + treatment:logS + treatment:RatioPoll + logS:R.score + logS:S.score + RatioPoll:C.score + R.score:C.score + C.score:S.score

The results show that the factor treatment (grazing) by itself has no significant effect (p-value = 0.21 > 0.05) on the height of plants and same goes for R.score and S.score. 
This means that the effect of treatment that we previously saw was mainly due to interactions between treatment and other factors.


