---
title: "Effets du pâturage sur les prairies suédoises"
author: "Bastien, Elia, Caroline, Alexia, Marine, Lily, Julie" 
bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      INTRODUCTION
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Terrestrial ecosystems were dominated for several million years by large herbivorous mammals, which then declined at the end of the Pleistocene under the pressure of the first prehistoric human populations. The loss of these large herbivore species (aurochs, wild horses and anes, woolly rhinoceros...) was also synonymous with the loss of the ecosystem services they provided. @lundgren2020human

For several years, the “rewilding” movement has been aiming to restore these ecosystem services by reintroducing species into large European natural areas, and the present study seeks to assess the extent to which introduced herbivore species have an influence on vegetation. @garido2019experimental

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      MATERIALS AND METHODS
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


The study site is located approximately 17 km south of Uppsala, Sweden. On this study site, three plots were used. Each plot was divided in two, with an ungrazed part and a grazed part. Four young Gotland stallions were placed in each grazing area for three years.

Numerous variables were studied:
- plant height at maturity
- seed mass
- species richness
- plot area occupied by anemogamous plants 
- plot area occupied by entomogamous plants 
- RCS score *.
Other information was noted at each survey, such as the season, year and observer who carried out the inventory. Information on the insects found on the plot was also recorded, but will not be used in our analysis.

* The RCS score corresponds to an index of the strategy used by the plants. @Wood2018microbiomes

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      RESEARCH QUESTIONS
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

We have chosen to use the  @garido2019experimental dataset to answer three different questions. Firstly, does grazing have an effect on the plant species richness of plots? Secondly, does grazing have an effect on plant morphology? And finally, does grazing have an effect on plant pollination strategies?

To answer these questions, we transformed the initial dataset to obtain a dataset of 404 records. The variables used are :
- treatment (grazed or ungrazed)
- season
- year
- cover.area.bee (plot area occupied by entomogamous plants)
- cover.area.gene (plot area occupied by anemogamous plants)
- logh (log of plant height)
- logS (log of seed mass)
- richness (species richness)
- observer
- R.score
- S.score
- C.score
- ratio poll (cover.area.bee/cover.area.gene)

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      DATA IMPORT
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


```{r}
df <- read.table('pollen.csv', sep = ';',stringsAsFactors=TRUE, header = TRUE)
df$year <- as.factor(df$year)
summary (df)
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                      PACKAGES IMPORT
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


```{r}

library(tidyverse)
library(FactoMineR)
library(dplyr)
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
                               PART 1: Effect of grazing on species richness
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Data visualization
```{r}
ggplot(data=df)+
  geom_boxplot(aes(x=treatment, y=num.species))+
  geom_jitter(aes(x=treatment, y=num.species)) + 
  labs(x="Treatment", y="Species richness") +
  ggtitle('Effect of grassland on species richness')
```
The specific richness of grazed areas appears to be greater than in ungrazed areas. 

```{r}
hyp1<-lm(num.species~1,data=df)
hyp0<-lm(num.species~treatment,data=df)
anova(hyp0,hyp1)
```
The p-value is less than 0.05, so grazing has a significant effect on specific richness. 

# Choice of model: in this section, we assume that grazing (treatment) depends on the season and year of observation

```{r}
modComplet<-LinearModel(num.species~(treatment+season+year)*(treatment),data=df,selection = "aic")
modComplet

```
The results for the full model are as follows: p-value: 2.554e-16, AIC = 649, BIC = 681. However, the model chosen according to the AIC criterion shows the following results: p-value: 1.41e-17, AIC = 645.2 and BIC = 669.2. As the AIC criterion is smaller for this model than for the full model, we choose to retain: num.species ~ treatment + season + year + treatment:year. There is therefore an interaction between the pasture and its year of observation. 

# Validation of the chosen model with the Bartlett test
```{r}
modelAIC<-lm(num.species~treatment+season+year+treatment:year,data=df)

res<-residuals(modelAIC)
boxplot(res~treatment,data=df)
bartlett.test(res~treatment,data=df)

boxplot(res~season,data=df)
bartlett.test(res~season,data=df)

boxplot(res~year,data=df)
bartlett.test(res~year,data=df)

boxplot(res~treatment:year,data=df)
df$interaction <- interaction(df$treatment, df$year)
bartlett.test(res ~ interaction, data = df)
```
For the Bartlett tests, the p-values are all greater than 0.05, which means that homoscedasticity is respected for all our parameters, or that the variances are all equal for all our observations.

# Residual normality test with Shapiro
hist(res,main="Histogramme residus",xlab="Residus")
shapiro.test(res)

Residual normality is respected, as our p-value is greater than 0.05: the data are compatible with a normal distribution. Our linear model is therefore valid, as residual normality is a key assumption for applying this model.

# Conclusion part 1 on species richness
Species richness therefore depends on grazing, season and year, as well as on the interaction grazing:year.