---
editor: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

title: "Exploration données sur les manchots" group: "CHOULETTE Klervi,
DENIS Manon, DUPORT Maureen, MOTTE Arthur, POYARD Antoine, SESIA Marion,
TANGUY Céline"

bibliography: references.bib execute: freeze: auto output:
html_document: toc: true toc_float: true ---

```{r}
library(tidyverse)
library(ggplot2)
```

## Présentation du jeu de données :

**Columns name :**

-   studyName :Sampling expedition from which data were collected,
    generated, etc.\
-   Sample Number : an integer denoting the continuous numbering
    sequence for each sample\
-   Species : a character string denoting the penguin species\
-   Region : a character string denoting the region of Palmer LTER
    sampling grid\
-   Island : a character string denoting the island near Palmer Station
    where samples were collected\
-   Stage : a character string denoting reproductive stage at sampling\
-   Individual ID : a character string denoting the unique ID for each
    individual in dataset\
-   Clutch Completion : a character string denoting if the study nest
    observed with a full clutch, i.e., 2 eggs\
-   Date Egg : a date denoting the date study nest observed with 1 egg
    (sampled)\
-   Culmen Length : a number denoting the length of the dorsal ridge of
    a bird's bill (millimeters)\
-   Culmen Depth : a number denoting the depth of the dorsal ridge of a
    bird's bill (millimeters)\
-   Flipper Length : an integer denoting the length penguin flipper
    (millimeters)\
-   Body Mass : an integer denoting the penguin body mass (grams)\
-   Sex : a character string denoting the sex of an animal\
-   Delta 15 N : a number denoting the measure of the ratio of stable
    isotopes 15N:14N\
-   Delta 13 C : a number denoting the measure of the ratio of stable
    isotopes 13C:12C\
-   Comments : a character string with text providing additional
    relevant information for data

```{r}
data <- read.table("penguins_raw.csv", sep = "," , header = TRUE, stringsAsFactors = T)
summary(data)
```

In this Data set, we would like to compare the different qualitative
variables :\
- *Culmen Length*\
- *Culmen Depth*\
- *Flipper Length*\
- *Body Mass*\
- *Delta 13C*\
- *Delta 15N*\

```{r}
sub_data_num = data%>%
  select(Culmen.Length..mm.,Culmen.Depth..mm.,Flipper.Length..mm.,Body.Mass..g.,Delta.13.C..o.oo.,Delta.15.N..o.oo.)%>%
  na.omit() # we delete the few individuals with NA

matcor = cor(sub_data_num,method = "pearson")
matcor
```

The representation of this matrix isn't the best. We can use the library
*corrplot* to make some visual correlation matrix.

We used the following code:

```{r}
library(corrplot)

corrplot(matcor, method = "number", type = "upper",order = "hclust", tl.col="black", tl.srt=45)


corrplot(matcor, type = "upper",order = "hclust", tl.col="black", tl.srt=45)

```

Now, we want to observe whether certain qualitative variables have an
effect on body mass. We will therefore try to create a graph for each
factor (qualitative variables) according to body mass:\
- *Species*\
- *Island*\
- *Sex*\

```{r}
box1 <- ggplot(data = data)+
  geom_boxplot(aes(x=Species,y=Body.Mass..g., fill = Species))
box1

box2 <- ggplot(data = data)+
  geom_boxplot(aes(x=Island,y=Body.Mass..g., fill = Island))
box2

box3 <- ggplot(data = data)+
  geom_boxplot(aes(x=Sex,y=Body.Mass..g., fill = Sex))
box3
```

We have noticed some trends that we will have to confirm with
statistical tests.\
Box1: Gentoo penguins appear to be heavier than the other two species.\
Box2: penguins from Biscoe Island appear to be heavier than the other
two islands.\
Box3: male penguins appear to be heavier than the other two categories.\
We run anovas to confirm these differences.\

```{r}
mod1= lm(Body.Mass..g.~Species, data = data)

anova1 = anova(mod1)

anova1

mod2 = lm(Body.Mass..g.~Island, data = data)

anova2 = anova(mod2)

anova2

mod3 = lm(Body.Mass..g.~Sex, data = data)

anova3 = anova(mod3)

anova3
```

Significance tests confirm the trends seen by the eye.
