---
title: "Exploring data on penguins" 
author: "CHOULETTE Klervi, DENIS Manon, DUPORT Maureen, MOTTE Arthur, POYARD Antoine, SESIA Marion, TANGUY Céline" 
bibliography: references.bib 
execute:
  freeze: auto
  warning: false
output: 
  html_document: 
    toc: true 
    toc_float: true 
    toc_depth: 3
---

# Introduction

## Study context

This document is a report on the analysis of a dataset on penguins. This dataset contains information on various [variables](#column_name). The dataset was find on kaggle : https://www.kaggle.com/datasets/utkarshx27/penguin-size-clutch-and-blood-isotope-data?select=penguins_raw.csv

This dataset was made from differents articles METTRE BIBLIO. Only 3 different species were studied : Adelie, Chinstrap and Gentoo. The dataset contains 344 observations and 17 variables. The goal of this analysis is to explore the dataset and to identify the main characteristics of the penguins.

This document is a report on the analysis of a dataset on penguins. This dataset contains information on various [variables](#colum-name). This dataset was made from differents articles Selon Gorman et al. [@10.1371/journal.pone.0090081] and Horst et al. [@10.32614/RJ-2022-020]. Only 3 different species where studied : Adelie, Chinstrap and Gentoo. The dataset contains 344 observations and 17 variables. The goal of this analysis is to explore the dataset and to identify the main characteristics of the penguins.

```{r, results='hide'}
#| message: false
#| warning: false
#| echo: true
#| results: hide
#| code-fold: true
#| code-summary: "Show the code"


library(factoextra) # for data visualization
library(FactoMineR) # for PCA
library(tidyverse)  # for data manipulation and visualization
library(dplyr)      # for data manipulation
library(ggplot2)    # for data visualization
library(corrplot)   # for correlation test

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(gridExtra)
library(kableExtra)
```

## Location

Data were collected in Antarctica on Biscoe, Dream and Torgersen Islands.These three blue points show the location of the data collect in @fig-map_antarctica.

```{r}
#| label: fig-map_antarctica
#| fig-cap: "Maps of Antarctica with the collect points"
#| code-fold: true
#| code-summary: "Show the code"
# Download country data (including Antarctica)
world <- ne_countries(scale = "medium", returnclass = "sf")

# Filter to keep only Antarctica
antarctica <- world[world$continent == "Antarctica", ]

rectangle_coords <- st_as_sfc(st_bbox(c(xmin = -90, xmax = -55, ymin = -75, ymax = -60), crs = st_crs(4326)))

antarctic_proj <- "+proj=laea +lat_0=-90 +datum=WGS84"
# Transform the rectangle into the polar projection system
rectangle_proj <- st_transform(rectangle_coords, crs = antarctic_proj)

# Create the map
gg_antartique = ggplot(data = antarctica) +
    annotate("rect", xmin = -90, xmax = -55, ymin = -75, ymax = -60,
           fill = "blue") +
  geom_sf() +
  coord_sf(crs = "+proj=laea +lat_0=-90 +datum=WGS84") +
 annotate("rect", 
           xmin = st_bbox(rectangle_proj)["xmin"], xmax = st_bbox(rectangle_proj)["xmax"], 
           ymin = st_bbox(rectangle_proj)["ymin"], ymax = st_bbox(rectangle_proj)["ymax"], 
           alpha = 0.2, fill = "blue")+
  theme_minimal() +
  labs(title = "Antarctica map",
       caption = "Data: Natural Earth") +
  annotation_scale(location = "bl", width_hint = 0.5)

antarctic_proj <- "+proj=laea +lat_0=-90 +datum=WGS84"

# Point coordinates in longitude and latitude
point_coords <- data.frame(lon = c(-66.209547, -64, -65), lat = c(-66, -65, -64.8))

# Convert the point into an sf object (longitude/latitude coordinate system WGS84)
point_sf <- st_as_sf(point_coords, coords = c("lon", "lat"), crs = 4326)

# Transform the point into the projected coordinate system of the map
point_proj <- st_transform(point_sf, crs = antarctic_proj)

# Extract the projected coordinates for use in annotate
point_proj_coords <- st_coordinates(point_proj)

# Create a bounding box manually with the coordinates
# Convert the bounding box into an sf object
bbox_sf <- st_as_sfc(st_bbox(c(xmin = -90, xmax = -55, ymin = -75, ymax = -60)))

# Crop the map using the bounding box
antarctica_cropped <- st_crop(antarctica, bbox_sf)

# Create the map
gg_zone_interet = ggplot() +
  geom_sf(data = antarctica_cropped) +
  geom_sf(data = point_proj, color = "blue", size = 3, alpha = 0.8, shape = 19) +

  coord_sf(crs = "+proj=laea +lat_0=-90 +datum=WGS84") +
  theme_minimal() +
  labs(title = "Area of Interest",
       caption = "Data: Natural Earth") +
  annotation_scale(location = "bl", width_hint = 0.5)

grid.arrange(gg_antartique, gg_zone_interet, ncol = 2)

```

## Dataset presentation {#column_name}

**Columns name :**

-   studyName : Sampling expedition from which data were collected, generated, etc.\
-   Sample Number : an integer denoting the continuous numbering sequence for each sample\
-   Species : a character string denoting the penguin species\
-   Region : a character string denoting the region of Palmer LTER sampling grid\
-   Island : a character string denoting the island near Palmer Station where samples were collected\
-   Stage : a character string denoting reproductive stage at sampling\
-   Individual ID : a character string denoting the unique ID for each individual in dataset\
-   Clutch Completion : a character string denoting if the study nest observed with a full clutch, i.e., 2 eggs\
-   Date Egg : a date denoting the date study nest observed with 1 egg (sampled)\
-   Culmen Length : a number denoting the length of the dorsal ridge of a bird's bill (millimeters)\
-   Culmen Depth : a number denoting the depth of the dorsal ridge of a bird's bill (millimeters)\
-   Flipper Length : an integer denoting the length penguin flipper (millimeters)\
-   Body Mass : an integer denoting the penguin body mass (grams)\
-   Sex : a character string denoting the sex of an animal\
-   Delta 15 N : a number denoting the measure of the ratio of stable isotopes 15N:14N\
-   Delta 13 C : a number denoting the measure of the ratio of stable isotopes 13C:12C\
-   Comments : a character string with text providing additional relevant information for data

Loading dataset :

```{r}
data <- read.table("penguins_raw.csv", sep = "," , header = TRUE, stringsAsFactors = T)
summary(data)
# dim(data)
```

Names modification of columns and species :

```{r}
#| code-fold: true
#| code-summary: "Show the code"
data <- data %>%
  select(-X) %>%                   # Remove X column 
  select(-Comments) %>%           # Remove Comments column
  rename(                          # we rename the columns of the dataset
    sample_number = Sample.Number,
    id = Individual.ID,
    island = Island,
    region = Region,
    stage = Stage,
    clutch_completion = Clutch.Completion,
    date_egg = Date.Egg,
    culmen_length = Culmen.Length..mm.,
    culmen_depth = Culmen.Depth..mm.,
    flipper_length = Flipper.Length..mm.,
    body_mass = Body.Mass..g.,
    sex = Sex,
    delta_15_N = Delta.15.N..o.oo.,
    delta_13_C = Delta.13.C..o.oo.,
    species = Species)

names(data)

# rename species
data = data %>%
  mutate(species = recode(species, "Adelie Penguin (Pygoscelis adeliae)" = "Adelie", "Chinstrap penguin (Pygoscelis antarctica)" = "Chinstrap", "Gentoo penguin (Pygoscelis papua)" = "Gentoo"))

```

These names are easier to understand.

## Missing values :

The @fig-missing-val have a red line that represents the mean of NA values in the dataset and the bars are showing the number of NA's for each variables.

```{r}
#| label: fig-missing-val
#| fig-cap: "Number of missing value as a function of each variables"
#| code-fold: true
#| code-summary: "Show the code"

# Verification of missing values
colSums(is.na(data)) 

data_NA <- data 

# Representation of missing values
NA_df <- (colSums(is.na(data_NA)) / nrow(data_NA)) * 100

color_pal <- colorRampPalette(c("limegreen", "yellow", "firebrick"), bias = 20)(10)
NA_col <- color_pal[as.integer(NA_df/length(NA_df)*9)+1]

# Graphique
barplot(NA_df,ylab = "NA", col = NA_col, las=2, cex.names=0.6)

NA_mean<-mean(NA_df)
# 0.78% 

abline(h=NA_mean, col = "red", lwd = 2)


```

# Exploratory analysis

In this Data set, we would like to compare the different qualitative variables :\
- *Culmen Length*\
- *Culmen Depth*\
- *Flipper Length*\
- *Body Mass*\
- *Delta 13C*\
- *Delta 15N*\

The representation of this matrix isn't the best.

```{r}
sub_data_num = data%>%
  select(culmen_length, culmen_depth,flipper_length,body_mass,delta_13_C,delta_15_N)%>%
  na.omit() # we delete the few individuals with NA

matcor = cor(sub_data_num,method = "pearson")
matcor
```

We can use the library *corrplot* to make some visual correlation matrix.

There is a correlation that we have to remove in a multivariate analysis between the Body Mass and the flipper length as we can see in @fig-cor_mat2 and @fig-cor_mat3.

```{r}
#| label: fig-cor_mat2
#| fig-cap: "Correlation matrix with numbers"
#| code-fold: true
#| code-summary: "Show the code"

corrplot(matcor, method = "number", type = "upper",order = "hclust", tl.col="black", tl.srt=45)
```

```{r}
#| label: fig-cor_mat3
#| fig-cap: "Correlation matrix with forms"
#| code-fold: true
#| code-summary: "Show the code"
corrplot(matcor, type = "upper",order = "hclust", tl.col="black", tl.srt=45)
```

Now, we want to observe whether certain qualitative variables have an effect on body mass. We will therefore try to create a graph for each factor (qualitative variables) according to body mass:\
- *Species*\
- *Island*\
- *Sex*\

We have noticed some trends in @fig-boxplot_trends that we will have to confirm with statistical tests.\
Box1: Gentoo penguins appear to be heavier than the other two species.\
Box2: Penguins from Biscoe Island appear to be heavier than the other two islands.\
Box3: Male penguins appear to be heavier than the other two categories.

```{r}
#| label: fig-boxplot_trends
#| fig-cap: "Boxplots between body mass and other variables"
#| code-fold: true
#| code-summary: "Show the code"

box1 <- ggplot(data = data)+
  geom_boxplot(aes(x=species,y=body_mass, fill = species))

box2 <- ggplot(data = data)+
  geom_boxplot(aes(x=island,y=body_mass, fill = island))

box3 <- ggplot(data = data)+
  geom_boxplot(aes(x=sex,y=body_mass, fill = sex))
global_box <- grid.arrange(box1,box2,box3)
```

## Violin plot :

Here is another way to visualize in @fig-violin the distribution between two variables.

```{r violin, fig.height=3, fig.width=9}
#| label: fig-violin
#| fig-cap: Violin plot of body mass as function of many variables
#| code-fold: true
#| code-summary: "Show the code"


# Violin plot for Culmen Length (beak length)
v1 <- ggplot(data, aes(x = species, y = culmen_length, fill = species)) +
  geom_violin(trim = FALSE) +
  geom_jitter(shape = 16, position = position_jitter(0.2), color = "black", alpha = 0.3) +
    stat_summary(fun = mean, geom = "point", shape = 20, size = 3, fill = "red", color = "red") +
  labs(title = "Distribution of Culmen Length by Species",
       x = "Species",
       y = "Culmen Length (mm)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 10, face = "bold"),
        axis.title.x = element_text(size = 8),  # Taille de l'axe x
        axis.title.y = element_text(size = 8))

# Violin plot for Flipper Length
v2 <- ggplot(data, aes(x = species, y = flipper_length, fill = species)) +
  geom_violin(trim = FALSE) +
  geom_jitter(shape = 16, position = position_jitter(0.2), color = "black", alpha = 0.3) +
    stat_summary(fun = mean, geom = "point", shape = 20, size = 3, fill = "red", color = "red") +
  labs(title = "Distribution of Flipper Length by Species",
       x = "Species",
       y = "Flipper Length (mm)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 10, face = "bold"),
        axis.title.x = element_text(size = 8),  # Taille de l'axe x
        axis.title.y = element_text(size = 8))

# Violin plot for Body Mass
v3 <- ggplot(data, aes(x = species, y = body_mass, fill = species)) +
  geom_violin(trim = FALSE) +
  geom_jitter(shape = 16, position = position_jitter(0.2), color = "black", alpha = 0.3) +
    stat_summary(fun = mean, geom = "point", shape = 20, size = 3, fill = "red", color = "red") +
  labs(title = "Distribution of Body Mass by Species",
       x = "Species",
       y = "Body Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(size = 10, face = "bold"),
        axis.title.x = element_text(size = 8),  # Taille de l'axe x
        axis.title.y = element_text(size = 8))

grid.arrange(v1, v2, v3,  #names of graphs  
             ncol = 3, nrow = 1)
```

We run anovas to confirm these differences.\
There is a significant **difference between the body mass of these 3 species**, almost 2 of them, with a p-value of *2e-16*.

```{r}
anova1 = aov(data$body_mass~data$species)

summary(anova1)
```

There is a significant **difference between the body mass of these 3 islands**, almost 2 of them with a p-value of *2e-16*.

```{r}
anova2 = aov(body_mass~island, data = data)

summary(anova2)
```

There is a significant **difference between the body mass and the sex** with a p-value of *4.9e-16*.

```{r}
anova3 = aov(body_mass~sex, data = data)

summary(anova3)
```

Distribution of species between the islands in @tbl-table-distribution : we can see from this table that the *Adelie* species is present on all three islands, while the *Chrinstrap* and *Gentoo* species are only present on 2 islands, Dream and Biscoe respectively.

```{r}
#| label: tbl-table-distribution
#| tbl-cap: Species distribution between the islands
#| code-fold: true
#| code-summary: "Show the code"

species_islands <- table(data$island, data$species)
kable(species_islands, "html", escape = F) %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                position = "center")
```

Some graphs and interpretations :

In @fig-bdm-lenght, it would appear that there is a positive linear trend between the body_mass variable and flipper_lenght for each specie.

```{r}
#| label: fig-bdm-lenght
#| fig-cap: Flipper length as a function of body mass 
#| code-fold: true
#| code-summary: "Show the code"

ggplot(data, aes(x = flipper_length, y = body_mass, color = species))+
  geom_point()+
  scale_color_manual(values = c("#B2DF8A","#FDBF6F","#CAB2D6"))+
  geom_smooth(method = "lm", col = "red", se = TRUE)+
  theme_bw()+
  theme(legend.position = "none")+
  facet_grid(~species)
```

Check with linear regression :

```{r}

rl1 <- lm(body_mass ~ flipper_length + species + species:flipper_length, data = data)
summary(rl1)

```

These results show a highly significant effect of fin length for all species, with a greater effect for the *Gentoo* species. There was also a significant difference in body mass between species except between *Adélie* and *Chinstrap*.

In conclusion, we can say that fin length influences body mass and that this effect varies according to species.

It would appear that there is a positive linear trend between the body_mass variable and flipper_lenght for each specie.

Check with linear regression :

```{r}

rl1 <- lm(body_mass ~ flipper_length + species + species:flipper_length, data = data)
summary(rl1)

```

These results show a highly significant effect of fin length for all species, with a greater effect for the *Gentoo* species. There was also a significant difference in body mass between species except between *Adélie* and *Chinstrap*.

In conclusion, we can say that fin length influences body mass and that this effect varies according to species.

In the @fig-lenght-depth, it would appear that there is a positive linear trend between the culmen_depth variable and culmen_length for each specie.

```{r}
#| label: fig-lenght-depth
#| fig-cap: Culmen length as a function of culmen depth 
#| code-fold: true
#| code-summary: "Show the code"

ggplot(data, aes(x = culmen_length, y = culmen_depth, color = species))+
  geom_point()+
  scale_color_manual(values = c("#B2DF8A","#FDBF6F","#CAB2D6"))+
  geom_smooth(method = "lm", col = "red", se = TRUE)+
  theme_bw()+
  theme(legend.position = "none")+
  facet_grid(~species)
```

Check with linear regression :

```{r}
rl2 <- lm(culmen_depth ~ culmen_length + species + species:culmen_length, data = data)
summary(rl2)
```

These results show that beak length is significant and positive in *Adélie*. In the species *Chinstrap* and *Gentoo*, the beak depth is smaller and is significant in *Gentoo*. At the interaction level, it is not significant, which means that the effect of bill length on depth is relatively the same for 3 species.

We can conclude that beak depth depends on beak length but this does not vary between species. Gentoo\* penguins have a shallower beak than the others.

In @fig-delta, in the There seems to be different trends between species.

```{r, fig.width=10, fig.height=5}
#| label: fig-delta
#| fig-cap: Delta_15_N as a function of delta_13_C 
#| code-fold: true
#| code-summary: "Show the code"

ggplot(data, aes(x = delta_15_N, y = delta_13_C, color = species))+
  geom_point()+
  scale_color_manual(values = c("#B2DF8A","#FDBF6F","#CAB2D6"))+
  geom_smooth(method = "lm", col = "red", se = TRUE)+
  theme_bw()+
  theme(legend.position = "none")+
  facet_grid(~species)
```

We check this using a linear model :

```{r}
rl3 <- lm(delta_13_C ~ delta_15_N + species + species:delta_15_N, data = data)
summary(rl3)
```

These results show that the relationship between nitrogen 15 and carbon 13 is positive and significant in *Adélie*, this relationship is reduced in *Chinstrap* and even more so in *Gentoo*. *Chinstrap* and *Gentoo* have significantly higher carbon-13 values than *Adélie*. The interactions between nitrogen 15 and the species are significant, showing that the effect of nitrogen 15 on carbon 13 varies according to the species.

In conclusion, this model shows that the effect of nitrogen 15 on carbon 13 is highly species-dependent, with differences between species.

## PCA on the species caracteristics

```{r}
# Selection of quantitative data
data_numeric <- select_if(data, is.numeric)

# Remove sample_number 
# The body_mass et the flipper_length are more than 80% correlated, we're deleting flipper_length because it's the data that's most correlated with all the other data.
# names(data_numeric)
data_numeric <- data_numeric %>% select(-c("sample_number","flipper_length"))

# Normalizing data
data_scales <- scale(data_numeric)

# PCA realization
acp_data= PCA(data_scales, graph = F)
summary(acp_data)
```

@fig-eig_value_plot is an histogram of the variance of the factorial axes with the percentage of inertia of each axis. Here, keep 2 dimensions and 78.4% inertia.

```{r}
#| label: fig-eig_value_plot
#| fig-cap: Barplot of eigen values
#| code-fold: true
#| code-summary: "Show the code"
#Variance for each component
va_propre = fviz_eig(acp_data, main ="", choice = "variance",
         label_size =50, addlabels = T,
         barfill  = "grey",barcolor = "grey")+
   ylim(c(0,60))
va_propre
```

@fig-contribution_plot determind the absolut contribution of variables to each axis. We can see that 3 variables make a strong contribution to axis 1 (delta_15_N, body_mass and culmen_depth) and 2 to axis 2 (culmen_depth and delta_13_C).

```{r,warning=FALSE,message=FALSE, fig.height=2, fig.width=5, fig.align = "center"}
#| label: fig-contribution_plot
#| fig-cap: Barplots of absolut contributions of axe 1 and 2
#| code-fold: true
#| code-summary: "Show the code"

#Contribution of axis variables axis 1 and 2
c1 = fviz_contrib(acp_data,choice = "var",axes =1,
                  fill = "grey",color = "grey")+ labs(title = "Axe 1") #For axe 1
c2 = fviz_contrib(acp_data,choice = "var",axes =2,
                  fill = "grey",color = "grey")+ labs(title = "Axe 2") #For axe 2
grid.arrange(c1,c2, ncol = 2)
```

First graph (@fig-projection) show the result of PCA on axes 1 and 2. *A priori* clustering by the Species has been performed. Gentoo penguins have a greater mass and a longer culmen. Adelies are lighter and have a depth culmen. Chinstrap penguins are the lightest and have a long, depth culmen. Adelies and Chinstrap penguins both have a high concentration of carbon 13 and nitrogen 15.

```{r,warning=FALSE,message=FALSE, fig.height=7, fig.width=12, fig.align = "center"}
#| label: fig-projection
#| fig-cap: Projection of variables and individuals according to species
#| code-fold: true
#| code-summary: "Show the code"

#Projection of variables and individuals and differentiation according to species
bi_plot = fviz_pca_biplot(acp_data, label="var", 
                habillage=data$species, 
                addEllipses = T,
                ellipse.alpha = F, 
                ylim=c(-3,3.5), xlim = c(-3.5,3.5))+
  scale_color_manual(values = c("#B2DF8A","#FDBF6F","#CAB2D6"))+
  theme_minimal()+
  theme(legend.position = "bottom")
grid.arrange(bi_plot, va_propre, c1, c2,  #names of graphs
             ncol = 2, nrow = 3, #How many row and col
             widths = c(2.2,1), #Here, the first col will be bigger than the second
             layout_matrix = cbind(c(1,1,1), #On the first col only graph 1 for the 3 rows
                                   c(2,3,4))) #On the second col, the 3 other graphs
```

Second graph (@fig-projection-islands) show the result of PCA on axes 1 and 2. *A priori* clustering by the Island has been performed. We can see that Gentoo penguins live only on Biscoe Island, while Chinstrap penguins live mainly on Dream Island. Adelie penguins live on these two islands and also on Togersen Island.

The common habitats of Chinstrap and Adelie penguins seem to explain the presence of isotopes in these two species, as they must find a similar diet.

```{r}
#| label: fig-projection-islands
#| fig-cap: Projection of variables and individuals and dfferentiation according to the islands
#| code-fold: true
#| code-summary: "Show the code"
#Projection of variables and individuals and differentiation according to species 
bi_plot = fviz_pca_biplot(acp_data, label="var",  habillage=data$island,  addEllipses = T,  ellipse.alpha = F,   ylim=c(-3,3.5),xlim = c(-3.5,3.5))+  
  scale_color_manual(values = c("#B2DF8A","#FDBF6F","#CAB2D6"))+  
  theme_minimal()+  theme(legend.position = "bottom") 
grid.arrange(bi_plot, va_propre, c1, c2, #names of graphs  
             ncol = 2, nrow = 3, #How many row and col  
             widths = c(2.2,1), #Here, the first col will be bigger than the second  
             layout_matrix = cbind(c(1,1,1), #On the first col only graph 1 for the 3 rows  
                                   c(2,3,4))) #On the second col, the 3 other graphs
```

There is a significant *difference between the body mass of these 3 species*, almost 2 of them.

```{r}
anova2 = aov(body_mass~island, data = data)

summary(anova2)
```

There is a significant *difference between the body mass of these 3 islands*, almost 2 of them.

```{r}
anova3 = aov(body_mass~sex, data = data)

summary(anova3)
```

There is a significant *difference between the body mass and the sex.*

# Body mass model: an example of ANCOVA

ANCOVA (i.e. analysis of variance-covariance) is a statistic method that enable to model the relation between a dependent quantitative variable and several dependent explanatory variable that can be **quantitative AND qualitative**.

Here we aim to determine:

1.  Which variable influence the penguins' body mass of the penguins

2.  What is the equation of the body mass model

We are selecting different variable of the data set that we think can influence the body mass: *species, island, clutch_completion, culmen_length, culmen_depth, flipper_length, body_mass, sex, delta_15_N, delta_13_C*. They are the explanatory variables. It is important to check the type (numeric or factor) of each variable and change if necessary.

We remove the individuals for which at least one measure is missing.

```{r data preparation}
#| results: hide


#selection of the interesting variables
dataLM<-data %>% 
  select(species,island,clutch_completion,culmen_length,culmen_depth,flipper_length,body_mass,sex,delta_15_N,delta_13_C)

#set the variable type (factors or numerics)
dataLM$species<-as.factor(dataLM$species)
dataLM$island<-as.factor(dataLM$island)
dataLM$clutch_completion<-as.factor(dataLM$clutch_completion)
dataLM$sex<-as.factor(dataLM$sex)
dataLM$flipper_length<-as.numeric(dataLM$flipper_length)
dataLM$body_mass<-as.numeric(dataLM$body_mass)

str(dataLM)

#check if there is missing values 
colSums(is.na(dataLM))

#delete individual with missing values
dataLM<-dataLM %>% 
  drop_na(culmen_length,culmen_depth,flipper_length,body_mass,sex,delta_15_N,delta_13_C)
```

## Data exploration

Before any modelisation, we have to explore the data to avoid some mistake. These are the different step to follow prior to execute a model:

1.  Check for aberrant values and distribution shape of the response variable which is here *body_mass*.
2.  If the explanatory variable is **quantitative**, check for aberrant values and distribution shape, and if the explanatory variable is **qualitative**, analyse the number mode and the number of individual in each one.
3.  Analyse the potential relation between the response variable and the different explanatory variables.
4.  Look for potential interaction between explanatory variables.
5.  Check for potential colinearity between explanatory variables.

### 1. Response variable exploration

```{r, fig.height=5, fig.width=15}
#| label: fig-lm_resp_varia
#| fig-cap: Distribution of the values of the variable body mass
#| code-fold: true
#| code-summary: "Show the code"

par(mfrow=c(1,3))
# Cleveland plot
dotchart(dataLM$body_mass,pch=16,col='dodgerblue3',xlab='Body mass')
#Histogram
hist(dataLM$body_mass,col='dodgerblue3',xlab="body mass",main="")
# Quantile-Quantile plot
qqnorm(dataLM$body_mass,pch=16,col='dodgerblue3',xlab='')
qqline(dataLM$body_mass,col='red',lwd=2)
```

We can see on the first graphic of the @fig-lm_resp_varia that there is no aberrant values for the variable body mass. The two other graphic of the figure show that there is a deviation of normality. However, normality of the response variable is not a requirement for ANCOVA.

### 2. Explanatory variables exploration

First we invastigate the distribution of the quantitative explanatory variables.

```{r covariable, include=TRUE, fig.height=12, fig.width=9}
#| label: fig-lm_expl_varia
#| fig-cap: Distribution of the values of the explanatory variables
#| code-fold: true
#| code-summary: "Show the code"


par(mfrow=c(5,3))

# Culmen Length
# Cleveland plot
dotchart(dataLM$culmen_length,pch=16,col='dodgerblue3',xlab='culmen_length')
# Histogram
hist(dataLM$culmen_length,col='dodgerblue3',xlab="culmen_length",main="")
# Quantile-Quantile plot
qqnorm(dataLM$culmen_length,pch=16,col='dodgerblue3',xlab='')
qqline(dataLM$culmen_length,col='red',lwd=2)

# Culmen Depth
# Cleveland plot
dotchart(dataLM$culmen_depth,pch=16,col='dodgerblue3',xlab='culmen_depth')
# Histogram
hist(dataLM$culmen_depth,col='dodgerblue3',xlab="culmen_depth",main="")
# Quantile-Quantile plot
qqnorm(dataLM$culmen_depth,pch=16,col='dodgerblue3',xlab='')
qqline(dataLM$culmen_depth,col='red',lwd=2)

# Flipper length
# Cleveland plot
dotchart(dataLM$flipper_length,pch=16,col='dodgerblue3',xlab='flipper length')
# Histogram
hist(dataLM$flipper_length,col='dodgerblue3',xlab="flipper length",main="")
# Quantile-Quantile plot
qqnorm(dataLM$flipper_length,pch=16,col='dodgerblue3',xlab='')
qqline(dataLM$flipper_length,col='red',lwd=2)

# Delta_15_N
# Cleveland plot
dotchart(dataLM$delta_15_N,pch=16,col='dodgerblue3',xlab='delta_15_N')
# Histogram
hist(dataLM$delta_15_N,col='dodgerblue3',xlab="delta_15_N",main="")
# Quantile-Quantile plot
qqnorm(dataLM$delta_15_N,pch=16,col='dodgerblue3',xlab='')
qqline(dataLM$delta_15_N,col='red',lwd=2)

# Delta_13_C
# Cleveland plot
dotchart(dataLM$delta_13_C,pch=16,col='dodgerblue3',xlab='delta_13_C')
# Histogram
hist(dataLM$delta_13_C,col='dodgerblue3',xlab="delta_13_C",main="")
# Quantile-Quantile plot
qqnorm(dataLM$delta_13_C,pch=16,col='dodgerblue3',xlab='')
qqline(dataLM$delta_13_C,col='red',lwd=2)
```

We can see on the first graphic of each row of the @fig-lm_expl_varia that there is no abberant values for the variable *culmen_length*, *culmen_depth*, *flipper_length*, *delta_15_N* and *delat_13_C*. The two other graphic of each row of the figure show that there is a deviation of normality for all explanatory variable. However, normality of the explanatory variable is not a requirement for ANCOVA.

Secondly we analyse the qualitative explanatory variables. We look at the distribution of the values in the different mode.

```{r cofactors, include=TRUE}
# Factor Sex
summary(dataLM$sex)
# Factor Island
summary(dataLM$island)
# Factor Species
summary(dataLM$species)
# Factor Clutch completion
summary(dataLM$clutch_completion)
```

We observe that there is not the same number of observation in each mode so the plan is non orthogonal. This is not mandatory but it is best when the plan is orthogonal.

### 3. Analysis of potential relation between response and explanatories variables

We explore the dynamics between body mass and the explanatory variable, to see if there is some trend.

```{r datagraph, include=TRUE, fig.height=15, fig.width=6}
#| label: fig-lm_mass_effect
#| fig-cap: Effect of the explanatories variables on the body mass
#| code-fold: true
#| code-summary: "Show the code"

par(mfrow=c(5,2))
# Culmen length
plot(dataLM$body_mass~dataLM$culmen_length,pch=16,col='dodgerblue3',xlab='culmen length',ylab='body mass')
# Culmen depth
plot(dataLM$body_mass~dataLM$culmen_depth,pch=16,col='dodgerblue3',xlab='culmen depth',ylab='body mass')
# Flipper length
plot(dataLM$body_mass~dataLM$flipper_length,pch=16,col='dodgerblue3',xlab='flipper length',ylab='body mass')
# delta 15 N
plot(dataLM$body_mass~dataLM$delta_15_N,pch=16,col='dodgerblue3',xlab='delta 15N',ylab='body mass')
# delta 13 C
plot(dataLM$body_mass~dataLM$delta_13_C,pch=16,col='dodgerblue3',xlab='delta 13C',ylab='body mass')
#Species
boxplot(dataLM$body_mass~dataLM$species, varwidth = TRUE, ylab = "body mass", xlab = "Species", col='darkgrey', main = "")
#Island
boxplot(dataLM$body_mass~dataLM$island, varwidth = TRUE, ylab = "body mass", xlab = "Island", col='brown', main = "")
#Clutch completion
boxplot(dataLM$body_mass~dataLM$clutch_completion, varwidth = TRUE, ylab = "body mass", xlab = "Clutch completion", col='brown', main = "")
#Sex
boxplot(dataLM$body_mass~dataLM$sex, varwidth = TRUE, ylab = "body mass", xlab = "Sex", col='brown', main = "")
```

We can see on the @fig-lm_mass_effect that the variables *culmen length, culmen depth, flipper length, delta 15N, species* and *island* seems to influence the body mass. The body mass seems to increase when *culmen length, culmen depth* and *flipper length* increase. On the opposite, body mass seems to decrease when *delta 15N* increase. The species Gentoo and the individuals of the island Biscoe seem to have a higher body mass.

### 4. Look for potential interaction between explanatory variables

Here we are considering a simple model with no interaction but in some case it can usefull to examine the potential interaction between the different variables.

### 5. Check of potential colinearity bewteen explanatories variables

In order to avoid colinearity in modelling, we analyse how the explanatories variables are related. To do so we calculate the correlation between the quantitative explanatories variables and evaluate if the qualitative explanatory variables impact the quantitative one using boxplot.

First between covariable: the quantitative explanatories variables.

```{r, fig.height=5, fig.width=5}
#| label: fig-lm_colinearity_covariable
#| fig-cap: Corelation matrice between the covariables
#| code-fold: true
#| code-summary: "Show the code"


corrplot.mixed(cor(dataLM %>% select_if(is.numeric)%>%select(-body_mass)),upper="square",lower.col="black", tl.col="black",cl.cex = 0.8,tl.cex = 0.7,number.cex =0.8)
```

The threshold above which we cosider there is colinearity is usually choose at +/- 0.7. We can thus see in the @fig-lm_colinearity_covariable that there is no colinearity detected between the covariables.

Then between cofactors: the qualitative explanatory variables.

```{r colinearity cofactors,include=TRUE, fig.height=15, fig.width=12}
#| label: fig-lm_colinearity_cofactor
#| fig-cap: Exploration of potential colinearities between cofactors and covariable
#| code-fold: true
#| code-summary: "Show the code"


# Checking collinearity between cofactors and covariables
par(mfrow=c(5,4))
#Culmen length and species
boxplot(dataLM$culmen_length~dataLM$species, varwidth = TRUE, ylab = "Culmen length", xlab = "Species", col='grey', main = "")
#Culmen length and island
boxplot(dataLM$culmen_length~dataLM$island, varwidth = TRUE, ylab = "Culmen length", xlab = "Island", col='grey', main = "")
#Culmen length and sex
boxplot(dataLM$culmen_length~dataLM$sex, varwidth = TRUE, ylab = "Culmen length", xlab = "Sex", col='grey', main = "")
#Culmen length and clutch completion
boxplot(dataLM$culmen_length~dataLM$clutch_completion, varwidth = TRUE, ylab = "Culmen length", xlab = "Clutch completion", col='grey', main = "")

#Culmen depth and species
boxplot(dataLM$culmen_depth~dataLM$species, varwidth = TRUE, ylab = "Culmen depth", xlab = "Species", col='grey', main = "")
#Culmen depth and island
boxplot(dataLM$culmen_depth~dataLM$island, varwidth = TRUE, ylab = "Culmen depth", xlab = "Island", col='grey', main = "")
#Culmen depth and sex
boxplot(dataLM$culmen_depth~dataLM$sex, varwidth = TRUE, ylab = "Culmen depth", xlab = "Sex", col='grey', main = "")
#Culmen depth and clutch completion
boxplot(dataLM$culmen_depth~dataLM$clutch_completion, varwidth = TRUE, ylab = "Culmen depth", xlab = "Clutch completion", col='grey', main = "")

#Flipper length and species
boxplot(dataLM$flipper_length~dataLM$species, varwidth = TRUE, ylab = "Flipper length", xlab = "Species", col='grey', main = "")
#Flipper length and island
boxplot(dataLM$flipper_length~dataLM$island, varwidth = TRUE, ylab = "Flipper length", xlab = "Island", col='grey', main = "")
#Flipper length and sex
boxplot(dataLM$flipper_length~dataLM$sex, varwidth = TRUE, ylab = "Flipper length", xlab = "Sex", col='grey', main = "")
#Flipper length and clutch completion
boxplot(dataLM$flipper_length~dataLM$clutch_completion, varwidth = TRUE, ylab = "Flipper length", xlab = "Clutch completion", col='grey', main = "")

#delta 15N and species
boxplot(dataLM$delta_15_N~dataLM$species, varwidth = TRUE, ylab = "delta 15N", xlab = "Species", col='grey', main = "")
#delta 15N and island
boxplot(dataLM$delta_15_N~dataLM$island, varwidth = TRUE, ylab = "delta 15N", xlab = "Island", col='grey', main = "")
#delta 15N and sex
boxplot(dataLM$delta_15_N~dataLM$sex, varwidth = TRUE, ylab = "delta 15N", xlab = "Sex", col='grey', main = "")
#delta 15N and clutch completion
boxplot(dataLM$delta_15_N~dataLM$clutch_completion, varwidth = TRUE, ylab = "delta 15N", xlab = "Clutch completion", col='grey', main = "")

#delta 13C and species
boxplot(dataLM$delta_13_C~dataLM$species, varwidth = TRUE, ylab = "delta 13C", xlab = "Species", col='grey', main = "")
#delta 13C and island
boxplot(dataLM$delta_13_C~dataLM$island, varwidth = TRUE, ylab = "delta 13C", xlab = "Island", col='grey', main = "")
#delta 13C and sex
boxplot(dataLM$delta_13_C~dataLM$sex, varwidth = TRUE, ylab = "delta 13C", xlab = "Sex", col='grey', main = "")
#delta 13C and clutch completion
boxplot(dataLM$delta_13_C~dataLM$clutch_completion, varwidth = TRUE, ylab = "delta 13C", xlab = "Clutch completion", col='grey', main = "")

```

According to the @fig-lm_colinearity_cofactor there is no colinearity between cofactors and covariables.

We keep all our explanatory variables to do the modelisation.

## Statistics analysis

For the statistic modelisation, we analyse first the full model, which means the model including all explanatories variables.

In order to identify the candidate model from the full model, we are going to use the backward selection model. This method is based on the significance of the model terms. From the full model, we identify the significance of each term of the model and proceed to a skimming of the non-sgnificant terms. The withdraw on the non significant term have to follow two step: - first step : remove successively the non significant interaction - second step : remove successively the explanatory variable that are non significant and not include in interaction terms.

First we build the full model:

```{r full model}
modlm<-lm(body_mass ~ species + island + clutch_completion + culmen_length + culmen_depth + flipper_length + sex + delta_15_N + delta_13_C,data=dataLM) #full model
```

Using the function *drop1* we analyse the significance of each variables:

```{r full model selection}
drop1(modlm,test="F") #test significativté des variables avec test fischer
```

We exclude the variable *island* of the model, because it is the non-significant variable with the highest p-value. Then, we create a new model (without the variable *island*) and test for significance:

```{r model2}
modlm2<-lm(body_mass ~ species + clutch_completion + culmen_length + 
    culmen_depth + flipper_length + sex + delta_15_N + delta_13_C, data=dataLM)
drop1(modlm2,test="F")
```

for the same reason as previously, we exclude the variable *clutch_completion* of the model. We build a new model and test again for variables' significance:

```{r model3}
modlm3<-lm(body_mass ~ species + culmen_length + 
    culmen_depth + flipper_length + sex + delta_15_N + delta_13_C, data=dataLM)
drop1(modlm3,test="F")
```

All the variable in the model are significant, so we keep all of them in our final model. The best model is thus: $body\:mass$ \~ $species + culmen_{length} + culmen_{depth} + flipper_{length} + sex + delta\:15\:N + delta\:13\:C$

Using the function *summary* we can acces the coefficient for all variable.

```{r}
summary(modlm3)
```

The model can thus be written as:

$$ Body\:Mass = 1467.1\:+\:(Species_{Adeliae} = 0 \:;\:Species_{Pygoscelis antarctica} = -245.1^{**}\:;\:Species_{Pygoscelis papua} = 882.1^{***})$$ $$ +\:17.1^{*NS*}. Culmen\:Length\:+\:61.6^{**}. Culmen\:Depth\:+\:17.8^{***}. Flipper\:Length $$ $$ + (Sex_{Female}= 0\:;\: Sex_{Male}=417.9^{***})\:-\:164.3^{***}. Delta\:15N\:+\:65.2^{*}. Delta\:13C)$$

From the summary of the model we can also access the variance proportion that is explained by the model. It correspond to the variable of *Adjusted R-squared* which is here 0.8813. The model is thus explaining 88% of the body mass variance.

## Model validation

The hypotheses for ANCOVA are **independence**, **residuals normality** and **variance homogeneity**. We also need to check for influent statistics unities, which means an observation with a high contribution to model construction.

Here, we assume an independence of variables as we don't have the details of the data collection.

We check the normality of residuals.

```{r ResidNorm, include=TRUE, fig.height=3, fig.width=6}
#| label: fig-lm_resid_norm
#| fig-cap: Verification of residuals normality
#| code-fold: true
#| code-summary: "Show the code"

par(mfrow=c(1,2))
# Histogram
hist(modlm3$residuals,col='dodgerblue3',xlab="residuals",main="Check Normality")
# Quantile-Quantile plot
qqnorm(modlm3$residuals,pch=16,col='dodgerblue3',xlab='')
qqline(modlm3$residuals,col='red',lwd=2)
```

We can see on the first plot of @fig-lm_resid_norm that the histogram seems to have a normal distribution. This is confirmed by the second plot of the figure where we can see that the residuals are align with the red line, which means that the residuals have a normal distribution.

We check the homoscedasticity of residuals.

```{r Residhomo, include=TRUE, fig.height=5, fig.width=5}
#| label: fig-lm_variance
#| fig-cap: verification of residuals homoscedasticity
#| code-fold: true
#| code-summary: "Show the code"


# residuals vs fitted
plot(residuals(modlm3)~fitted(modlm3)
      , col='dodgerblue3'
      , pch=16)
abline(h = 0)

```

We can see on the @fig-lm_variance that there is no tendency on the point repartition. We can thus conclude that the variance is homogeneous.

We also verify that there isn't an individual that is particularly influential.

```{r Contri, include=TRUE, fig.height=4, fig.width=4}
#| label: fig-lm_influent_indiv
#| fig-cap: Verification of influential individual
#| code-fold: true
#| code-summary: "Show the code"


par(mfrow = c(1, 1))
plot(cooks.distance(modlm3), type = "h", ylim = c(0, 1))
abline(h = 1, col = 2,lwd = 3)

```

As all the bar are lower that 1 on the @fig-lm_influent_indiv, we can conclude that there is no influential individual.

Finally, we can conclude that the **linear model is valid**.
