---

title: "Exploration données sur les manchots"
group: "CHOULETTE Klervi, DENIS Manon, DUPORT Maureen, MOTTE Arthur, POYARD Antoine, SESIA Marion, TANGUY Céline" 

bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---

## Study context

This document is a report on the analysis of a dataset on penguins. This dataset contains information on various [variables](#colum-name). This dataset was made from 4 differents articles METTRE BIBLIO.
Only 3 different species where studied : Adelie, Chinstrap and Gentoo. The dataset contains 344 observations and 17 variables. The goal of this analysis is to explore the dataset and to identify the main characteristics of the penguins.
AJOUTER LES QUESTIONS DE RECHERCHE ET LES ANALYSES FAITES


```{r}
library(tidyverse)
library(ggplot2)
```


## Présentation du jeu de données {#colum-name} : 

**Columns name :**

 - studyName :Sampling expedition from which data were collected, generated, etc.  
 - Sample Number : an integer denoting the continuous numbering sequence for each sample  
 - Species : a character string denoting the penguin species  
 - Region : a character string denoting the region of Palmer LTER sampling grid  
 - Island : a character string denoting the island near Palmer Station where samples were collected  
 - Stage : a character string denoting reproductive stage at sampling  
 - Individual ID : a character string denoting the unique ID for each individual in dataset  
 - Clutch Completion : a character string denoting if the study nest observed with a full clutch, i.e., 2 eggs  
 - Date Egg : a date denoting the date study nest observed with 1 egg (sampled)  
 - Culmen Length : a number denoting the length of the dorsal ridge of a bird's bill (millimeters)  
 - Culmen Depth : a number denoting the depth of the dorsal ridge of a bird's bill (millimeters)  
 -  Flipper Length : an integer denoting the length penguin flipper (millimeters)  
 - Body Mass : an integer denoting the penguin body mass (grams)  
 - Sex : a character string denoting the sex of an animal  
 -  Delta 15 N : a number denoting the measure of the ratio of stable isotopes 15N:14N  
 - Delta 13 C : a number denoting the measure of the ratio of stable isotopes 13C:12C  
 - Comments : a character string with text providing additional relevant information for data  


```{r}
data <- read.table("penguins_raw.csv", sep = "," , header = TRUE, stringsAsFactors = T)
summary(data)
```



A citation @bauer2023writing
