---
title: "Exploration données sur les manchots"
group: "CHOULETTE Klervi, DENIS Manon, DUPORT Maureen, MOTTE Arthur, POYARD Antoine, SESIA Marion, TANGUY Céline" 

bibliography: references.bib
execute: 
  freeze: auto
output: 
  html_document:
   toc: true
   toc_float: true
---


```{r}
library(tidyverse)
library(ggplot2)
```


## Présentation du jeu de données : 

**Columns name :**

 - studyName :Sampling expedition from which data were collected, generated, etc.  
 - Sample Number : an integer denoting the continuous numbering sequence for each sample  
 - Species : a character string denoting the penguin species  
 - Region : a character string denoting the region of Palmer LTER sampling grid  
 - Island : a character string denoting the island near Palmer Station where samples were collected  
 - Stage : a character string denoting reproductive stage at sampling  
 - Individual ID : a character string denoting the unique ID for each individual in dataset  
 - Clutch Completion : a character string denoting if the study nest observed with a full clutch, i.e., 2 eggs  
 - Date Egg : a date denoting the date study nest observed with 1 egg (sampled)  
 - Culmen Length : a number denoting the length of the dorsal ridge of a bird's bill (millimeters)  
 - Culmen Depth : a number denoting the depth of the dorsal ridge of a bird's bill (millimeters)  
 -  Flipper Length : an integer denoting the length penguin flipper (millimeters)  
 - Body Mass : an integer denoting the penguin body mass (grams)  
 - Sex : a character string denoting the sex of an animal  
 -  Delta 15 N : a number denoting the measure of the ratio of stable isotopes 15N:14N  
 - Delta 13 C : a number denoting the measure of the ratio of stable isotopes 13C:12C  
 - Comments : a character string with text providing additional relevant information for data  


```{r}
data <- read.table("penguins_raw.csv", sep = "," , header = TRUE, stringsAsFactors = T)
summary(data)
```



